Red Wine Quality Analysis by Jake Williams
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggcorrplot)
library(ggplot2)
library(reshape2)
library(RColorBrewer)

```

```{r echo=FALSE, Load_the_Data}

# Set Working Directory and Load the Data
rwine <- read.csv('wineQualityReds.csv')

```

This report is going to be exploring a dataset containing 13 different 
variables in red wine. We will be analyzing the data to figure out what
defines '**quality**'.

### DISCLAIMER:

This dataset is public available for research. The details are described
in [Cortez et al., 2009]. 

  P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

[@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016
[Pre-press (pdf)] http://www3.dsi.uminho.pt/pcortez/winequality09.pdf
[bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib


# Univariate Plots Section

To begin, I want to get the structure of the data, meaning I want to see what
makes up our dataset as a whole.  

```{r echo=FALSE, Univariate_Plots}

str(rwine) #Getting the structure of the data

```

By looking at the summary above, our dataset contains 13 variables that we will
explore further. We can quickly notice that there is one variable that is only
serving as an index, variable 'X'.  This will be omitted from our studies and 
now we have 12 variables left to explore.  

Let's get an overview of these variables by looking at each as a histogram.
This will give us a better understanding of how red wines are distributed 
across our dataset. Our X axis will display each variable while the Y axis
will be our number of wines.

```{r echo=FALSE}

stack(summary(rwine$fixed.acidity)) # mean, median, mode, max, etc about data of the variable

ggplot(rwine, aes(x=fixed.acidity)) + # data being used for x axis
geom_histogram(binwidth = .1, color = 'black', fill = 'dark green') + # histogram, bar width, outline color, fill color
labs(title = 'Fixed Acidity Distribution', x='Fixed Acidity Level(tartaric acid - g / dm^3)',
y = 'Number of Wines') + # adding title, x label, y label
scale_x_continuous(breaks = seq(0, 20, 1)) + # appropriate scaling of x ticks to better understand distribution
scale_y_continuous(breaks = seq(0, 100, 5)) # appropriate scaling of y ticks to better understand distribution

```

Our first graph displays the fixed acidity distribution in our dataset.  Most
of the red wine appears to fall between the 6.5 and 8.5 range.

```{r echo=FALSE}

stack(summary(rwine$volatile.acidity)) 

ggplot(rwine, aes(x=volatile.acidity)) +
geom_histogram(binwidth = .05, color = 'black', fill = 'springgreen4') + 
labs(title = 'Volatile Acidity Distribution', x='Volatile Acidity Level(acetic acid - g / dm^3)', 
y = 'Number of Wines') +
scale_x_continuous(breaks = seq(0, 2, .1)) + 
scale_y_continuous(breaks = seq(0, 200, 20))

```

Volatile acidity is measured in lower amounts  with a majority of wines
between .3 and .7.

```{r echo=FALSE}

stack(summary(rwine$citric.acid)) 

ggplot(rwine, aes(x=citric.acid)) + 
geom_histogram(binwidth = .01, color = 'black', fill = 'orange1') + 
labs(title = 'Citric Acid Distribution', x='Citric Acid Level(g / dm^3)', 
y = 'Number of Wines') + 
scale_x_continuous(breaks = seq(0, 2, .05)) +
scale_y_continuous(breaks = seq(0, 150, 10))

```

Citric acid levels are interesting.  Over 130 wines have close to a 0 citric
acid level while another larger portion of wines has closer to .5.

```{r echo=FALSE}

stack(summary(rwine$residual.sugar)) 

ggplot(rwine, aes(x=residual.sugar)) + 
geom_histogram(binwidth = .5, color = 'black', fill = 'cyan4') + 
labs(title = 'Residual Sugar Distribution', x='Residual Sugar Level(g / dm^3)', 
y = 'Number of Wines') + 
scale_x_continuous(breaks = seq(0, 16, 1)) +
scale_y_continuous(breaks = seq(0, 800, 50))

```

Red wines with low residual sugar levels are much more common than higher
levels.  Over 650 wines alone stay around 2.

```{r echo=FALSE}

stack(summary(rwine$chlorides)) 

ggplot(rwine, aes(x=chlorides)) + 
geom_histogram(binwidth = .01, color = 'black', fill = 'steelblue2') + 
labs(title = 'Chlorides Distribution', 
x='Chloride Level(sodium chloride - g / dm^3)', 
y = 'Number of Wines') + 
scale_x_continuous(breaks = seq(0, 1, .05)) +
scale_y_continuous(breaks = seq(0, 800, 50))

```

Similar to residual sugar, the level of chlorides in red wine are staying 
more at the lower end.  I wonder if a higher chloride level affects taste.

```{r echo=FALSE}

stack(summary(rwine$free.sulfur.dioxide)) 

ggplot(rwine, aes(x=free.sulfur.dioxide)) + 
geom_histogram(binwidth = 1, color = 'black', fill = 'purple') + 
labs(title = 'Free Sulfur Dioxide (SO2) Distribution', 
x='SO2 Level(mg / dm^3)', 
y = 'Number of Wines') + 
scale_x_continuous(breaks = seq(0, 80, 2)) +
scale_y_continuous(breaks = seq(0, 200, 10))

```

SO2 distribution appears to have a more of a 'consistent' drop in wines as the
levels increase, with a few exceptions here and there.

```{r echo=FALSE}

stack(summary(rwine$total.sulfur.dioxide)) 

ggplot(rwine, aes(x=total.sulfur.dioxide)) + 
geom_histogram(binwidth = 4, color = 'black', fill = 'firebrick2') + 
labs(title = 'Total Sulfur Dioxide (TSO2) Distribution', 
x='TSO2 Level(mg / dm^3)',
y = 'Number of Wines') + 
scale_x_continuous(breaks = seq(0, 300, 20)) +
scale_y_continuous(breaks = seq(0, 200, 10))

```

Most red wine in our dataset TSO2 levels between 0 and 50.  We can also note
some outliers after a large gap in the data.

```{r echo=FALSE}

stack(summary(rwine$density)) 

ggplot(rwine, aes(x=density)) + 
geom_histogram(binwidth = .0001, color = 'black', fill = 'blue') + 
labs(title = 'Density Distribution', x='Density(g / cm^3)', 
y = 'Number of Wines') + 
scale_x_continuous(breaks = seq(.99, 1.5, .002)) +
scale_y_continuous(breaks = seq(0, 100, 5))

```

Density is interesting as a majority of the wines in our dataset fall between
.995 and .998 levels.  This graph depicts more of a normal distribution.

```{r echo=FALSE}

stack(summary(rwine$pH)) 

ggplot(rwine, aes(x=pH)) + 
geom_histogram(binwidth = .03, color = 'black', fill = 'darkturquoise') + 
labs(title = 'pH Level Distribution', x='pH Level', y = 'Number of Wines') + 
scale_x_continuous(breaks = seq(0, 5, .2)) +
scale_y_continuous(breaks = seq(0, 200, 10))

```

pH level also shows a more normal distribution with red wine. 3.3-3.4 seems
to be the sweet spot for most wines.

```{r echo=FALSE}

stack(summary(rwine$sulphates)) 

ggplot(rwine, aes(x=sulphates)) + 
geom_histogram(binwidth = .04, color = 'black', fill = 'darkred') + 
labs(title = 'Sulphates Level Distribution', 
x='Sulphates Level(potassium sulphate - g / dm3)', 
y = 'Number of Wines') + 
scale_x_continuous(breaks = seq(0,2, .1)) +
scale_y_continuous(breaks = seq(0, 300, 20))

```

Lower level sulphates appear to be much more common than higher levels, with
less than 20 going past 1.5. Is higher quality more common with lower levels?

```{r echo=FALSE}

stack(summary(rwine$alcohol)) 

ggplot(rwine, aes(x=alcohol)) + 
geom_histogram(binwidth = .2, color = 'black', fill = 'goldenrod') + 
labs(title = 'Alcohol Level Distribution', x='Alcohol Level(% by volume)', 
y = 'Number of Wines') + 
scale_x_continuous(breaks = seq(0,20, .5)) +
scale_y_continuous(breaks = seq(0, 500, 20))

```

I was interested to see how alcohol level is distributed for red wine. Right
under a 9.5 alcohol level is most common with almost double the amount of wines
as the next highest distribution.

```{r echo=FALSE}

stack(summary(rwine$quality)) 

ggplot(rwine, aes(x=quality)) + 
geom_histogram(binwidth = .5, color = 'black', fill = 'darkorchid4') + 
labs(title = 'Quality Distribution', x='Quality Level(between 0 and 10)', 
y = 'Number of Wines') + 
scale_x_continuous(breaks = seq(0,10, 1)) +
scale_y_continuous(breaks = seq(0, 1000, 100))

```

Finally, our quality distribution in the dataset.  A majority of red wine falls
around 5 and 6 with very few past a 7 rating.  Judging wine quality must be 
very strict.

# Univariate Analysis

### What is the structure of your dataset?

There are 1599 red wines in the dataset with 12 variables(fixed acidity, 
volatile acidity, citric acid, residual sugar, chlorides, free sulfur dioxide
total sulfur dioxide, density , pH, sulphates, alcohol, quality) that 
determine the quality of red wine. 'X' is considered a 13th variable in the
dataset but only serves the purpose of indexing. Many of the variables deal
with smaller decimals as their values. 

### What is/are the main feature(s) of interest in your dataset?

Quality is the main feature in this dataset.  I want to take a deeper dive into 
what determines how a wine is rated.

### What do you think will help support your investigation?

I was first wanting looking at alcohol level and residual sugar, but I want to 
view the correlation of each variable to see what will play a role in how 
quality is measured. A majority of people like sugar and alcohol, so I feel 
like they will play a large part in red wine quality.

### Did you create any new variables from existing variables in the dataset?

No new variables were created from existing variables.

### Any unsual distributions or adjustments to the form of the data?

Over half of the graphs had some outliers in the data such as volatile acidity,
citric acid, residual sugar, chlorides, free sulfur dioxide, total sulfur 
dioxide, and sulphates.  Other graphs appeared to have outliers but there 
weren't as many as those mentioned.

A few variables also seemed to be 'long tailed' such as residual sugar, 
chlorides and free sulfur dioxide.

For each of these operations, I adjusted each of their axis ranges as well as
bin width to better understand the distributions of the variable's data. Even
though these histograms are considered simplistic, one could put more variables
in lists depending on how the data's form was manipulated/plotted.

I also added color to the graphs to give it more of an easier to read visual.

# Bivariate Plots Section

A better visual for relationships in the data can be shown by viewing 
correlation levels.  Based on the graph, I want to explore variables 
that have a strong correlation with one another.

```{r echo=FALSE, Bivariate_Plots}

corr <- round(cor(rwine), 1) # store correlation of 'rwine' in 'corr'

ggcorrplot(corr, lab = TRUE) # plotting correlation

```

A strong correlation is defined by a higher positive number on the scale, and 
is represented by the color red. The darker the shade of red, the higher the 
correlation. To read this graph, we simply take one variable from the x axis 
and one variable from the y axis then find where the two intersect.

Correlation values have been rounded to provide an easier to read format.

Based on the visualization, some of the strongest correlation with variables 
are:

**Density** and **Fixed Acidity** (.7) \
**Citric Acid** and **Fixed Acidity** (.7) \
**Total Sulfur Dioxide(TSO2)** and **Free Sulfur Dioxide(SO2)** (.7) \

While quality seems to have the strongest correlation with:

**Citric Acid** (.2) \
**Sulphates** (.3) \
**Alcohol** (.5)  \

While some of the weakest correlation is:

**pH** and **Fixed Acidity** (-.7) \
**Citric Acid** and **Volatile Acidity** (-.6) \
**Alcohol** and **Density** (-.5) \


Let's look at our strong variable correlations and visualize each as a scatter
plot.

```{r echo=FALSE}

ggplot(rwine, aes(x=fixed.acidity, y=density)) +
geom_point(size=2, color = 'green') +
labs(title = 'Fixed Acidity and Density Correlation', x='Fixed Acidity', 
y = 'Density') +
scale_x_continuous(breaks = seq(0,16, 1)) +
scale_y_continuous(breaks = seq(.99, 2, .002))

```

We can clearly see the positive correlation (.7) here with a main focus on 
acidity levels being between 7 and 8 while density appears to be around the
.996 level.  

```{r echo=FALSE}

ggplot(rwine, aes(x=fixed.acidity, y=citric.acid)) +
geom_point(size=2, color = 'orange') +
labs(title = 'Fixed Acidity and Citric Acid Correlation', x='Fixed Acidity', 
y = 'Citric Acid') +
scale_x_continuous(breaks = seq(0,16, 1)) +
scale_y_continuous(breaks = seq(0, 2, .05))

```

Comparing fixed acidity and citric acid levels show another positive 
correlation(.7). Little to no citric acid levels (<.05) appear to favor fixed
acidity levels betwen 5 and 9.

```{r echo=FALSE}

ggplot(rwine, aes(x=total.sulfur.dioxide, y=free.sulfur.dioxide)) +
geom_point(size=2, color = 'blue') +
labs(title = 'SO2 and TSO2 Correlation', x='TSO2', y = 'SO2') +
scale_x_continuous(breaks = seq(0,300, 20)) +
scale_y_continuous(breaks = seq(0, 100, 10))

```

Both total sulfur dioxide and free sulfur dioxide appear to have very strong 
postive correlation(.7) with the scatter plot being in more of an upwards cone
shape. Both values increase together. Most of the correlation does focus on 
lower levels of both sulfur levels with a few oddities.

```{r echo=FALSE}

ggplot(rwine, aes(x=pH, y=fixed.acidity)) +
geom_point(size=2, color = 'darkred') +
labs(title = 'pH and Fixed Acidity Correlation', x='pH', y = 'Fixed Acidity') +
scale_x_continuous(breaks = seq(0,5, .2)) +
scale_y_continuous(breaks = seq(0, 20, 1))

```

I want  to look at a negatively correlated pair, so I chose pH and fixed 
acidity. The negative correlation(-.7) is clearly defined as there is a
noticable downward slope, and shows that as pH levels increase, fixed
acidity levels decrease.


```{r echo=FALSE}

ggplot(rwine, aes(x=alcohol, y=residual.sugar)) +
geom_point(size=2, color = 'purple4') +
labs(title = 'Alcohol and Residual Sugar Correlation', x='Alcohol', 
y = 'SO2') +
scale_x_continuous(breaks = seq(0,20, 1)) +
scale_y_continuous(breaks = seq(0, 20, 1))

```

Since I believed that alcohol and residual sugar affected quality, I was 
curious to see if there was any correlation between the two. By 
referencing our chart above, there was no correlation(0) due to no indication
of an upwards or downwards slope. We can visually see this, however that 
doesn't necessarily mean the variables on their own don't affect quality.


```{r echo=FALSE}

ggplot(rwine, aes(x=quality, y=residual.sugar)) +
geom_jitter(size =1, color = 'blue') +
labs(title = 'Residual Sugar and Quality Correlation', x='Quality',
y = 'Residual Sugar') +
scale_x_continuous(breaks = seq(0,10, .5)) +
scale_y_continuous(breaks = seq(0, 20, 1))

```

I thought that residual sugar levels would have an effect on the quality of a
red wine, however it appears that there is no correlation between them.  A 
majority ofwines are shown to have low residual sugar levels while still 
increasing in quality.

```{r echo=FALSE}

ggplot(rwine, aes(x=quality, y=alcohol)) +
geom_jitter(size =1, color = 'darkgreen') +
labs(title = 'Alcohol and Quality Correlation', x='Quality', y = 'Alcohol') +
scale_x_continuous(breaks = seq(0,10, .5)) +
scale_y_continuous(breaks = seq(0, 20, 1))
```

On the flip side, it came to no surprise(to me at least) that alcohol had an 
effect on quality.  We can see that while alcohol level increases, quality 
does increase more consistently with higher alcohol levels.  

```{r echo=FALSE}

ggplot(rwine, aes(x=quality, y=sulphates)) +
geom_jitter(size =1, color = 'purple') +
labs(title = 'Sulphates and Quality Correlation', x='Quality', 
y = 'Sulphates') +
scale_x_continuous(breaks = seq(0,10, .5)) +
scale_y_continuous(breaks = seq(0,2, .1))
```

Sulphates have a slight positive correlation with quality.  While most red 
wines are rated between 4.5 and 6.5, you can still notice a slight trend in the
data.

```{r echo=FALSE}

ggplot(rwine, aes(x=quality, y=citric.acid)) +
geom_jitter(size =1, color = 'red') +
labs(title = 'Citric Acid and Quality Correlation', x='Quality', 
y = 'Citric Acid') +
scale_x_continuous(breaks = seq(0,10, .5)) +
scale_y_continuous(breaks = seq(0,2, .1))

```

Citric acid is also shown to have a positive correlation with quality.


### Relationships observed and features of interest

I observed three positive correlations in the dataset as well as one negative 
correlation and one with no correlation.  I wanted to get a visual idea as to
how the data is distributed between two variables.  Two of the positive 
correlations involved fixed acidity as a variable with density/citric acid as 
the other variable. Both shared similar distributions with citric acid 
appearing to have more values closer to zero.

### Did you observe any interesting relationships between the other features?

I was surprised to see that total sulfur dioxide and free sulfur dioxide shared
a positive correlation.  I was under the impression that one would have a 
negative impact on the other, similar to putting diesel in a gas engine and 
vice versa.

### What was the strongest relationship you found?

Density and fixed acidity had the strongest positive correlation because of how it
has more of a consistent incline to the data.

pH and fixed acidity had the strongest negative correlation as the trend depicted
a sharper decline.

The strongest relationship with quality was alcohol(not surprised).

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

ggplot(rwine, aes(x = sulphates, y = alcohol)) + 
geom_point(aes(color = factor(quality))) + # color key is quality score
scale_color_brewer(palette = "RdYlBu", name="Quality", direction=-1)
```

I decided to find out if both alcohol and sulphates had any correlation with the 
quality of wine since both were the highest correlated variables when separately
compared to wine quality. Different colors define the quality score with red 
being the highest and blue being the lowest. Based on the graph, it actually
appears that there is a slgith positive correlation with both alcohol and 
sulphates.A majority of red wines are rated between 5 and 7 and the colors
representing them have a slight incline to them. 

Since alcohol has the highest correlation with quality, I want to compare both
alcohol and citric acid to see if their interactions show a positive
correlation with quality.

```{r echo=FALSE}

ggplot(rwine, aes(x = citric.acid, y = alcohol)) + 
geom_point(aes(color = factor(quality))) + # color key is quality score
scale_color_brewer(palette = "RdYlBu", name="Quality", direction=-1)

```

The result of these two variables show that there doesn't seem to be any 
correlation between them.  Even though each variable is a positive 
correlation when separatelycompared to quality, they simply don't share
the same trend together. Most higher rated wines appear to have a higher 
alcohol level and look to stay on the lower end of volatile acidity.


```{r echo=FALSE}

ggplot(rwine, aes(x = volatile.acidity, y = alcohol)) + 
geom_point(aes(color = factor(quality))) +# color key is quality score
scale_color_brewer(palette = "RdYlBu", name="Quality", direction=-1)
```

For this final graph, I wanted to look at the highest correlated variable and
the lowest correlated variable together and see how quality is depicted. It 
appears that most mid rated wines fall between the volatile acidity levels of
.4 and .8, while alcohol hovers around 9 and 10.  

# Multivariate Analysis

Alcohol and sulphates appear to have a very small positive correlation 
between each other.  You can see that there is some trend with the increase
of quality, however the amount of sulphates mainly stays near the lower end
of the x axis, between .5 and 1.


### Were there any interesting or surprising interactions between features?

It was interesting to see that higher alcohol and citric acid levels together
didn't influence quality scores as much as alcohol and sulphates.


# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

ggplot(rwine, aes(x=citric.acid)) + 
geom_histogram(binwidth = .01, color = 'black', fill = 'orange1') + 
labs(title = 'Citric Acid Distribution', x='Citric Acid Level', 
y = 'Number of Wines') + 
scale_x_continuous(breaks = seq(0, 2, .05)) +
scale_y_continuous(breaks = seq(0, 150, 10))

```

### Description One

The first graph I wanted to bring back as a final plot is the Citric Acid
distribution of the red wine dataset.  What I find most interesting is
that most wines have close to a 0 citric acid level and drops to lower counts,
but when you notice the citric acid level around .5, the number of wines
shoots back up to almost 70.  The data shows us how unique each wine can be
and how many variations exist.

### Plot Two
```{r echo=FALSE, Plot_Two}

ggplot(rwine, aes(x=quality)) + 
geom_histogram(binwidth = 1, color = 'black', fill = 'darkorchid4') + 
labs(title = 'Quality Distribution', x='Quality Level', y = 'Number of Wines') + 
scale_x_continuous(breaks = seq(0,10, 1)) +
scale_y_continuous(breaks = seq(0, 1000, 100))

```

### Description Two

Quality level of red wines was also very interesting.  I was expecting to
see many more higher rated wines but a majority of the dataset hovered 
around a 5 and 6 rating.  It appears that most wines out there are considered
"average".  

### Plot Three
```{r echo=FALSE, Plot_Three}

ggplot(rwine, aes(x=total.sulfur.dioxide, y=free.sulfur.dioxide)) +
geom_point(size=2, color = 'blue') +
labs(title = 'SO2 and TSO2 Correlation', x='TSO2', y = 'SO2') +
scale_x_continuous(breaks = seq(0,300, 20)) +
scale_y_continuous(breaks = seq(0, 100, 10))

```

### Description Three

The final graph I thought was interesting was the correlation between free
sulfur dioxide and total sulfur dioxide. I was not expecting for both to have
a positive correlation because in my mind, I see that one type would cause a
negative impact on the other, similar to using gasoline in a diesel car.


# Reflection

Coming into this datset I had no idea what kind of variables determine how a 
red wine(or any wine) is rated.  I do not consume alcohol so I wouldn't know
what a good wine even tastes like.  I probably would think they're all the 
same.

One struggle I ran into was the small amount of variables that affected 
my main focus(quality).  They were so similar in correlation that not many
of them stood out.  On the flip side, comparing other variables with each other
were a little easier due to their higher correlations.  

Regarding the dataset, I believe that a much larger dataset would achieve 
betterresults.  Since a majority of the dataset hovered around a 5 and 6
quality rating, I would've liked to see more wines that were of a much better 
quality.

Overall, the most surprising thing about this dataset was the amount of 
variables that go into making a quality wine.  I can't believe how scientific
the research gets and now understand that wine making is simply an art form
and should be respected.

